---
title: "Computational Biology Course Summary SoSe21"
author: Finn Lo
output:
  html_document:
    toc: yes
    number_section: no
    fig_caption: true
  pdf_document:
    toc: yes
  always_allow_html: true
date: "`r format(Sys.time(), '%d %B %Y')`"
---

# CHAPTER 5 - BRANDT ET AL. 2019 - Cy_TOF NORMAL

## 5.1 Loading libraries and data

we will analyse Cy-TOF data from mouse intestinal organoids
``` {r}
library(tidyverse)
library(viridis)

data      <- read_csv("~/Documents/Programming/R/BioInf/Dateien für Kurs-20210609/limos_180907_R_course.csv")
name.key  <- read_csv("~/Documents/Programming/R/BioInf/Dateien für Kurs-20210609/name_key.csv")
```

``` {r}
#### what are the columns? Type, range, values?
arrange(data)
data %>%
  dplyr::select(Ce140Di:Pt195Di) %>%
  summary()
arrange(name.key)
#### which perturbations are contained in the data?
#### names of cell lines?
```

``` {r}
#### Option 1, renaming:
col_table <- tibble(colnames = colnames(data)) %>%
  left_join(name.key, by = c("colnames" = "channel")) %>%
  mutate(new_colnames = ifelse(is.na(name), colnames, name))
colnames(data) <- col_table %>%
  pull(new_colnames)

#### Option 2:
col_rename <- tibble(channel = colnames(data)) %>%
  left_join(name.key) %>%
  mutate(new_colnames = ifelse(is.na(name), channel, name)) %>%
  pull(new_colnames)
new_columns <- name.key$channel
names(new_columns) <- name.key$name
data <- data %>%
  dplyr::rename(new_columns)

```

## 5.2 Quality Checks
``` {r}
#### plot scatter plots of one channel vs. time --> to spot irregularities
#### alpha < 1 makes density more visible
#### adding density info (--> geom_density_2d) helps interpretation

data %>%
  ggplot(aes(Time, DNA_191Ir)) + #col = line
  geom_point(alpha = 0.2) +
  geom_density_2d() +
  geom_smooth()

```

``` {r}
library(hexbin)
data %>%
  ggplot(aes(Time, DNA_191Ir)) +
  geom_hex(bins = 128,
           show.legend = F, aes(col = ..count..)) +
  scale_fill_viridis(option = "inferno") +
  scale_color_viridis(option = "inferno") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank()
  )
```

``` {r}
# plot 4 markers in facet_wrap
data_piv <- data %>%
  dplyr::slice(1:10000) %>%
  pivot_longer(values_to = "value",
               names_to = "marker",
               cols = c(Axin2, cCasp3, Ce140Di, CD24)) %>%
  dplyr::select(Time, value, marker)

  
data_piv %>%
  ggplot(aes(Time, value)) +
  geom_point(alpha = 0.3, size = 0.5) +
  geom_density_2d() +
  geom_smooth() +
  facet_wrap(~ marker, scales = "free_y")
  
```

## 5.3 Filter and Scale the data
``` {r}
#### produce scatterplot
data %>%
  ggplot(aes(Ce140Di, DNA_191Ir)) +
  geom_point(size = 0.1)
```

``` {r}
#### set cutoff at ~ 750, blue is nice
data %>%
  ggplot(aes(Ce140Di, DNA_191Ir)) +
  geom_point(size = 0.1) +
  geom_vline(xintercept = 750, col = "blue")

```

``` {r}
#### create new column --> beads_out: T or F
# Normalization beads were measured alongside our cells. These beads are 
# labelled with a well defined mixture of isotopes across the entire mass range 
# and were used to account for day to day sensitivity changes in the machine’s detector. 
# We need to filter out the bead events, so they don’t end up in our analysis. 
# Because only beads are labelled with Cerium 140 (Ce140Di) we can use that channel 
# to set an easy 1D cut-off.
data <- data %>%
  mutate(beads_out = ifelse(Ce140Di > 750, F,T))

data %>%
  dplyr::count(., beads_out) %>%
  mutate(perc = n / sum(n) * 100)
```

``` {r}
#### look at values of the markers
#### visualize distributions with box-/violin plots
#### beads should be filtered out
data %>%
  pivot_longer(pChk2:Dead_cells_195Pt,
               names_to = "marker",
               values_to = "value") %>%
  ggplot(aes(marker, log2(value), col = marker)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

``` {r}
data %>%
  pivot_longer(pChk2:Dead_cells_195Pt,
               names_to = "marker",
               values_to = "value") #%>%
  ggplot(aes(marker, log2(value), col = marker)) +
  geom_violin() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

``` {r}
#### use pseudo-count to include "0" values
data <- data %>%
  mutate_at(vars(pChk2:Dead_cells_195Pt), function(x) log2(x+1))
```

``` {r}
#### Challenges in mass cytometry:
#### lack of scatter info =!= flow cytometry
#### therefore DNA-Intercalator & Event length used --> ggplot
data %>%
  filter(beads_out) %>%
  ggplot(aes(DNA_191Ir, Event_length)) +
  geom_point(alpha = 0.1, size = 0.3) +
  geom_density_2d() +
  geom_vline(xintercept = 7.25, col = "blue") +
  geom_vline(xintercept = 9.25, col = "blue") +
  geom_hline(yintercept = 12, col = "blue") +
  geom_hline(yintercept = 16, col = "blue")
```

``` {r}
#### visualize a new 2D filter that would filter out doublets and debris
data <- data %>%
  mutate(singlets = ifelse(  Event_length <= 16 &
                             Event_length >= 12 &
                             DNA_191Ir >= 7.25 &
                             DNA_191Ir <= 9.25, T, F))
  
#### detect apoptotic cells & cell debris
#### use the CISPLATIN stained samples (compound binds covalently to proteins,
#### but cannot easily penetrate the cell membrane)
#### ==> only cells with a comprimised cell membrane should have a high signal in the PLATIN channel!

#### make scatter plot of DNA vs. dead cells (no beads, singlets only)
#### new filter against high platin content (apop + debris)
data %>%
  filter(beads_out) %>%
  filter(singlets) %>%
  ggplot(aes(DNA_191Ir, Dead_cells_195Pt)) +
  geom_point(alpha = 0.3) +
  geom_density_2d() +
  geom_hline(yintercept = 11.25, col = "blue")
```

``` {r}
#### create new col called "low_platin" 
#### --> singlets get label "T"
#### --> doublets + debris get "F"
data <- data %>%
  mutate(low_platin = ifelse(Dead_cells_195Pt < 11.25, T,F))
```

``` {r}
#### scatter plot DNA vs. dead cells, color-coded by low_platin
#### no beads, singlets only

data %>%
  filter(beads_out) %>%
  filter(singlets) %>%
  ggplot(aes(DNA_191Ir, Dead_cells_195Pt, col = low_platin)) +
  geom_point(alpha = 0.3) +
  geom_density2d()
```

``` {r}
data %>%
  filter(beads_out) %>%
  filter(singlets) %>%
  filter(low_platin == T) %>%
  ggplot(aes(DNA_191Ir, Dead_cells_195Pt, col = low_platin)) +
  geom_point(alpha = 0.3) +
  geom_density2d()
#### Percentage of data with no beads + singlets + low_platin
#### 95483/114932 = 0.8307782 ??
```

## 5.4 Principal Component Analysis
``` {r}
pca.plot.data <- data %>%
  filter(  beads_out &
           singlets &
           low_platin)
#### my Alternative
#low_pt <- data %>%
#  filter(beads_out) %>%
#  filter(singlets) %>%
#  filter(low_platin == T)

#### optional to get plot like in paper: filter untreated cells too
#### filter(treatment == "untreated" & line == "KRAS")

pca <- data %>%
  filter(beads_out, singlets, low_platin) %>%
  dplyr::select(c(CD24, CD44, EphB2, Krt20, cCasp3)) %>%
  scale() %>%
  prcomp()

pca.plot.data <- pca$x %>%
  as_tibble() %>%
  bind_cols(pca.plot.data, .)

#### normal plot
pca.plot.data %>%
  ggplot(aes(PC1, PC2)) +
  #scale_x_reverse() +
  geom_point(size = 0.1)
```

``` {r}
#### colored by CD44
pca.plot.data %>%
  ggplot(aes(PC1, PC2)) +
  geom_point(size = 0.1) +
  scale_color_viridis(option = "inferno")
```

``` {r}
#### color the dots in your PCA using the values of different signalling markers
pca_col_data <- pca.plot.data %>%
  pivot_longer(values_to = "value",
               names_to = "marker",
               cols = c(cCasp3, CD24, CD44, EphB2, Krt20)) %>%
  dplyr::select(PC1, PC2, value, marker)

pca_col_data %>%
  ggplot(aes(PC1, PC2, col = value)) +
  geom_point(size = 0.1) +
  scale_color_viridis(option = "inferno") +
  facet_wrap(~ marker)

#### diff. color options
# inferno
# magma
# plasma
# inferno
# cividis
```

## 5.5 Clustering
``` {r}
#### k Means
#### cluster into 6 clustersm color code the pca plot by cluster
#### cluster cells multiple times, what happens?

fitk <- pca.plot.data %>%
  dplyr::select(c(CD24, CD44, EphB2, Krt20, cCasp3)) %>%
  scale() %>%
  kmeans(centers = 6)

pca.plot.data <- pca.plot.data %>%
  mutate(cluster = as.factor(fitk$cluster))

#library(RColorBrewer)
#display.brewer.all()
pca.plot.data %>%
  ggplot(aes(PC1, PC2, col = cluster)) +
  geom_point(size = 0.1) +
  scale_color_brewer(palette = "Dark2",
                     direction = -1)
```

## 5.6 Heatmap representation of perturbation data
``` {r}
#### visualize changes in signalling by cluster, treatment, cell line w/ HEATMAP
#### group by cluster, treatment, cell line
#### calculate mean signal values within groups for markers of choice
#### plot w/ geom_tile & facet_grid

heatmap_data <- pca.plot.data %>%
  pivot_longer(c("p4e-BP1",
                 "Axin2",
                 "pERK", 
                 "IkBa", 
                 "pMEK",
                 "p-p38",
                 "pS6"),
               names_to = "marker",
               values_to = "value") %>%
  group_by(cluster, treatment, line, marker) %>%
  summarise(value = mean(value))

heatmap_data %>%
  ggplot(aes(marker,treatment, fill = value)) +
  geom_tile() +
  facet_grid(cluster ~ line) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_viridis(option = "inferno")
#### there are some clusters with higher overall signalling activity
#### how are these clusters characterized? 
#### (how is the expression of the cell-type markers in these clusters???)
```

``` {r}
#### Better assess the effects of treatments:
#### calculate signal changes compared to the untreated control
#### plot results in similar heatmap
background <- pca.plot.data %>%
  pivot_longer(c("p4e-BP1",
                 "Axin2",
                 "pERK", 
                 "IkBa", 
                 "pMEK",
                 "p-p38",
                 "pS6"),
               names_to = "marker",
               values_to = "value") %>%
  #### (1) compute mean signal values for the control condition of FLUC untreated
  filter(treatment == "untreated" & line == "FLUC") %>%
  group_by(marker) %>%
  summarise(background_value = mean(value))

heatmap_data %>%
  left_join(background, by = "marker") %>%
  #### (2) substact these background values from all other conditions (log2)
  mutate(log2_fold_change = value - background_value) %>%
  #### (3) plot w/ geom_tile  +  facet_grid
  ggplot(aes(marker,treatment, fill = log2_fold_change)) +
  geom_tile() +
  facet_grid(cluster ~ line) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_gradient2(low = "blue",
                       high = "red",
                       midpoint = 0)
```






# CHAPTER 6 - Cy-TOF ADVANCED #################################################

## 6.1. Loading libraries and content
``` {r}
library(tidyverse)
library(cowplot)
library(viridis)
library(scales)
library(sp)
#install.packages("hexbin")
library(hexbin)
theme_set(theme_cowplot())

data      <- read_csv("~/Documents/Programming/R/BioInf/Dateien für Kurs-20210609/limos_180907_R_course.csv")
name.key  <- read_csv("~/Documents/Programming/R/BioInf/Dateien für Kurs-20210609/name_key.csv")# %>% deframe()
#data_named <- data %>% rename(all_of(name.key))

arrange(data)
arrange(name.key)
col_table <- tibble(colnames = colnames(data)) %>%
  left_join(name.key, by = c("colnames" = "channel")) %>%
  mutate(new_colnames = ifelse(is.na(name), colnames, name))
colnames(data) <- col_table %>%
  pull(new_colnames)
```

## 6.2 First Quality Checks (advanced), improved graphics
``` {r}
data %>%
  ggplot(aes(Time, DNA_191Ir)) +
  geom_hex(bins = 128,
           show.legend = F, aes(col = ..count..)) +
  scale_fill_viridis(option = "inferno") +
  scale_color_viridis(option = "inferno") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank()
        )
```

``` {r}
data %>%
  ggplot(aes(Ce140Di, DNA_191Ir)) +
  geom_point(size = 0.1) +
  geom_vline(xintercept = 750, col = "blue")
```

``` {r}
data <- data %>%
  mutate(beads_out = between(Ce140Di, -Inf, 750))
# normal: data <- data %>%
#  mutate(beads_out = ifelse(Ce140Di > 750, F,T))

plot <- data %>%
  pivot_longer(pChk2:Dead_cells_195Pt,
               names_to = "marker",
               values_to = "values") %>%
  ggplot(aes(marker, values, col = marker)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

``` {r}
plot + geom_boxplot()
```

``` {r}
plot + geom_violin()
```

``` {r}
plot + geom_violin() + scale_y_log10()
```

``` {r}
# change to log2 scale
plot +
  geom_violin() +
  scale_y_continuous(trans = "log1p")
```

``` {r}
# advanced solution: create a custom "asinh" transformer
## asinh gives obvious hyperbolic functions, respectively compute 
## hyperbolic cosine, sine, tangent, + inverses
asinh_trans <- function(x) {
  trans_new(
    name = "asinh",
    transform = "asinh",
    inverse = "sinh",
    format = trans_format("log10", math_format(10^.x))
  )
}

plot + geom_violin() + scale_y_continuous(trans = "asinh", breaks = c(10^(1:4)))
```

``` {r}
# plot as density plots instead, use facet_wrap to splot up multiple plots by marker
data %>%
  pivot_longer(pChk2:Dead_cells_195Pt,
               names_to = "marker",
               values_to = "value") %>%
  ggplot(aes(value, col = marker)) +
  geom_line(stat = "density") +
  facet_wrap(~ marker, scales = "free_y", ncol = 4) +
  scale_x_continuous(trans = "asinh", breaks = c(10^(1:4)))
```

``` {r}
# filter data for beads_out
# produce scatter plot for DNA vs. Event_length
# visualize a new 2D filter for filtering doublets + debris, save filter

data %>%
  filter(beads_out) %>%
  ggplot(aes(DNA_191Ir, Event_length)) +
  geom_point(size = 0.1) +
  geom_density_2d() +
  annotate("rect",
           xmin = 70, xmax = 1300,
           ymin = 12, ymax = 15,
           col = "red", alpha = 0) +
  scale_x_continuous(trans = "asinh",
                     breaks = c(10^(1:4)),
                     limits = c(0, 1e4)) +
  background_grid(major = "xy",
                  minor = "none")
```

``` {r}
data <- data %>%
  mutate(singlets = between(DNA_191Ir, 70, 1300) & 
                    between(Event_length, 12, 15))

#### detect apoptotic cells & cell debris
#### use the CISPLATIN stained samples (compound binds covalently to proteins,
#### but cannot easily penetrate the cell membrane)
#### ==> only cells with a comprimised cell membrane should have a high signal in the PLATIN channel!
#### make scatter plot of DNA vs. dead cells (no beads, singlets only)
#### new filter against high platin content (apop + debris)

data %>%
  filter(beads_out) %>%
  ggplot(aes(DNA_191Ir, Dead_cells_195Pt)) +
  geom_point(size = 0.1) +
  geom_density_2d() +
  background_grid(major = "xy", minor = "none") +
  scale_x_continuous(trans = "asinh", breaks = c(10^(1:4))) +
  scale_y_continuous(trans = "asinh", breaks = c(10^(1:4))) +
  # add polygon
  annotate(geom = "polygon",
           x = c(70, 70, 1300, 1300),
           y = c(8, 1000, 6000, 19),
           col = "blue",
           fill = "transparent")
```

``` {r}
data <- data %>%
  mutate(low_plat = point.in.polygon(point.x = DNA_191Ir,
                                     point.y = Dead_cells_195Pt,
                                     pol.x = c(70, 70, 1300, 1300),
                                     pol.y = c(8, 1000, 6000, 19)))

```

## 6.3 PCA Advanced
``` {r}
pca.plot.data <- data %>%
  dplyr::filter(beads_out & low_plat & singlets) #%>%
  filter(treatment == "untreated" & line == "KRAS")

pca <- pca.plot.data %>%
  dplyr::select(c(CD24, CD44, EphB2, Krt20, cCasp3)) %>%
  mutate_all(asinh) %>%
  scale() %>%
  prcomp()


pca.plot.data <- pca$x %>%
  as_tibble() %>%
  bind_cols(pca.plot.data, .)


pca.plot <- pca.plot.data %>%
  ggplot(aes(PC1, PC2)) +
  geom_point(size = 0.1) +
  scale_x_reverse()
```

``` {r}
p1 <- pca.plot +
  aes(col = pERK) +
  scale_color_viridis(option = "inferno",
                      trans = "asinh",
                      breaks = c(10^(1:4)))
p2 <- pca.plot +
  aes(col = pS6) +
  scale_color_viridis(option = "inferno",
                      trans = "asinh",
                      breaks = c(10^(1:4)))
p3 <- pca.plot +
  aes(col = EphB2) +
  scale_color_viridis(option = "inferno",
                      trans = "asinh",
                      breaks = c(10^(1:4)))
p4 <- pca.plot +
  aes(col = cCasp3) +
  scale_color_viridis(option = "inferno",
                      trans = "asinh",
                      breaks = c(10^(1:4)))
plot_grid(p1, p2, p3, p4, nrow = 2) # col options above: inferno, magma, plasma, cividis
```

``` {r}
plot_like_five <- pca.plot.data %>%
  pivot_longer(values_to = "value",
               names_to = "marker",
               cols = c(pERK, pS6, EphB2, cCasp3)) %>%
  dplyr::select(PC1, PC2, value, marker)


plot_like_five %>%
  ggplot(aes(PC1, PC2, col = value)) +
  geom_point(size = 0.1) +
  scale_x_reverse() +
  scale_color_viridis(option = "inferno",
                      trans = "asinh",
                      breaks = c(10^(1:4))) +
  facet_wrap(~ marker)
```

## 6.4 Clustering advanced (kMeans)
``` {r}
fitk <- pca.plot.data %>%
  dplyr::select(c(CD24, CD44, EphB2, Krt20, cCasp3)) %>%
  mutate_all(asinh) %>%
  kmeans(centers = 6)

pca.plot.data.fit <- pca.plot.data %>%
  mutate(cluster = as_factor(fitk$cluster))

pca.plot.data.fit %>%
  ggplot(aes(PC1, PC2, col = cluster)) +
  geom_point(size = 0.5)
```

## 6.5 Heatmap representation of perturbation data advanced
``` {r}
heatmap_data <- pca.plot.data.fit %>%
  pivot_longer(c("p4e-BP1", "Axin2", "pERK", "IkBa", "pMEK", "p-p38", "pS6"),
               names_to = "marker",
               values_to = "value") %>%
  group_by(cluster, treatment, line, marker) %>%
  summarise(value = mean(value))

heatmap_data %>%
  ggplot(aes(marker,treatment, fill = value)) +
  geom_tile() +
  facet_grid(cluster ~ line) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_viridis(option = "inferno", 
                     trans = "asinh", 
                     breaks = c(10^(1:4)))
```

``` {r}
# background stuff
background <- pca.plot.data.fit %>%
  pivot_longer(c("p4e-BP1", "Axin2", "pERK", "IkBa", "pMEK", "p-p38", "pS6"),
               names_to = "marker",
               values_to = "value") %>%
  mutate(value = log2(value + 1)) %>% # forgot this line.. now everything works!
  filter(treatment == "untreated", line == "FLUC") %>%
  group_by(marker) %>%
  summarise(background_value = mean(value))

```

## Bonus: Transform data points with log2(x + 1) to get more descriptive log2 fold changes
``` {r}
pca.plot.data.fit %>% 
  pivot_longer(c("p4e-BP1", "Axin2", "pERK", "IkBa", "pMEK", "p-p38", "pS6"), 
               names_to = "marker", 
               values_to = "value") %>%
  mutate(value = log2(value + 1)) %>% 
  group_by(cluster, treatment, line, marker) %>% 
  summarise(value = mean(value)) %>% 
  left_join(background, by = "marker") %>%
  mutate(log2fc = value - background_value) %>% 
  
  ggplot(aes(x = marker, y = treatment, fill = log2fc)) + 
  geom_tile() + facet_grid( cluster ~ line) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  scale_fill_gradient2(low = "blue",
                       high = "red", 
                       midpoint = 0)
```
