---
title: "Computational Biology Course Summary SoSe21"
author: Finn Lo
output:
  html_document:
    toc: yes
    number_section: no
    fig_caption: true
  pdf_document:
    toc: yes
  always_allow_html: true
date: "`r format(Sys.time(), '%d %B %Y')`"
---

```{r, echo = F, message = F, warning=F}
#setwd("/Users/finnLo/Documents/Github/R_Introduction/CompBio_Data/")
#R.version

library(ggplot2)
library(dplyr)
library(tidyr)
library(BiocManager)
library(bookdown)

# INTRO
library(BiocStyle)
library(Biostrings)
library(tidyverse)
library(janitor)
library(GEOquery)
library(biomaRt)
library(cgdsr)
library(ggdendro)
library(scatterplot3d)
```

# CHAPTER 1 - Data Analysis in R

Pipe shortcut mac Shift + Command + M

1.  Data loading
2.  Data formatting (tidying or harmonization)
3.  Data transformation
4.  Data modeling
5.  Data visualization
6.  Reporting

## 1.1. Data loading
```{r}
library(tidyverse)
library(stringr)
#### nested functions:
x1 <- 1:5
sqrt(
  sum(
    scale(
      x1, center = T, scale = F
    )^2
  )
)


#### chain of functions:
x1 %>%
  scale(center = T, scale = F) %>% 
  '^'(2) %>%
  sum() %>%
  sqrt()
```

## 1.1. Chains of Functions
``` {r}
x = "a, b, c;d:e-1"
x %>%
  str_replace_all(";", ":") %>%
  str_replace_all(":", ",") %>%
  str_remove("-1") %>%
  str_split(",", simplify = T)
```

## 1.2. Data Wrangling: loading and formatting
``` {r}
library(readr)
cdata <- read_tsv("https://raw.githubusercontent.com/tlobnow/Resources/main/R_Introduction/CompBio_Data/clinical_data.tsv")
cdata
```

```{r}
# Sex and gender columns are not harmonised

unique(cdata$SEX)
unique(cdata$GENDER)

# reformat columns
cdata <- cdata %>% 
  replace_na(list(SEX = "", GENDER = "")) %>% # replace all NA values by an empty string
  unite("SEX", SEX, GENDER, sep = "") %>%     # unite the SEX and GENDER column into a new SEX column
  mutate(SEX = toupper(SEX)) %>%              # convert all letters to uppercase
  mutate(SEX = ifelse(SEX == "FEMALE" | SEX == "MALE", SEX, "UNKNOWN")) 

ifelse(10 > 1, "stimmt", "falsch")
ifelse(10 < 1, "stimmt", "falsch")
ifelse(10 == 10 & 1 == 2, "stimmt", "falsch")
ifelse(10 == 10 | 1 == 2, "stimmt", "falsch")

```

## 1.3 Merging two data tables
``` {r}
#### left_join()
#### inner_join()
#### right_join()
#### bind_cols()
#### bind_rows()

aa <- tibble(aa = 1:3, bb = c("a", "b", "c"))
bb <- tibble(aa = 3:5, cc = c("d", "e", "f"))

aa %>% left_join(bb)
aa %>% right_join(bb)
aa %>% inner_join(bb)
aa %>% full_join(bb)
aa %>% bind_cols(bb)
aa %>% bind_rows(bb)
```

## 1.4 Data transformation
```{r}
#### select()
#### filter()
cdata
cdata %>% dplyr::select(STUDY_ID, CASE_ID)
cdata %>%dplyr::select(1:5)
cdata %>%dplyr::select(-STUDY_ID, -CASE_ID)
cdata %>%dplyr::select(-(1:5))
cdata %>%dplyr::select(contains("OS"))

cdata %>% filter(OS_STATUS == "DECEASED")
cdata %>% filter(SEX != "MALE")
cdata %>% filter(AGE > 50)
cdata %>% filter(AGE > 50, SEX == "FEMALE")
cdata %>% filter(OS_STATUS != "DECEASED", OS_STATUS != "LIVING")
#### OS_STATUS columns isn't all caps, some data is wrong

cdata <- cdata %>%
  mutate(OS_STATUS = toupper(OS_STATUS))
cdata %>% filter(OS_STATUS != "DECEASED", OS_STATUS != "LIVING")

cdata %>% 
  group_by(OS_STATUS, SEX) %>%
  tally()

cdata %>%
  group_by(STUDY_ID, SEX) %>%
  summarise(mean_age = mean(AGE, na.rm = T),
            sd_age = sd(AGE, na.rm = T)) %>%
  arrange(mean_age)

#### a) write statement with group_by() and summarise() that reproduced tally() data
#### b) write statement with group_by() and summarise() that calculates 
#### median overall survival grouped by sex and study
#### c) bonus: in what study do you find the lowest median value for patients of unknown sex

?n()

# a) 
cdata %>%
  group_by(OS_STATUS, SEX) %>%
  summarise(n())

# b) , c)
cdata %>%
  group_by(SEX, STUDY_ID) %>%
  summarise(median_overall_survival = median(AGE, na.rm = T)) %>%
  filter(SEX != "FEMALE", SEX != "MALE") %>%
  arrange(desc(median_overall_survival))

glimpse(cdata)
```

## 1.5 Data Visualization Examples
``` {r}
#### number of patients per study:
cdata %>% 
  group_by(STUDY_ID, SEX) %>% 
  tally %>% 
  ggplot(aes(STUDY_ID, n)) +
  geom_bar(stat = "identity") +
  coord_flip()
```

```{r}
#### Histogram of overall Survival:
ggplot(cdata, aes(OS_MONTHS)) +
  geom_histogram()
```

```{r}
#### Histogram of age in a subset of studies, colored by sex:
filter(cdata, str_detect(STUDY_ID, "brca|gbm|coadread")) %>% 
  ggplot(aes(AGE, fill = SEX)) +
  geom_histogram(position = "dodge") +
  facet_wrap(~STUDY_ID)
```

```{r}
#### Density plot of overall survival, grouped by disease free survival status:
filter(cdata, OS_STATUS == "LIVING") %>% 
  ggplot(aes(OS_MONTHS, color = DFS_STATUS)) +
  geom_density()
```

```{r}
#### Scatter Plot pf age and overall survival values
ggplot(cdata, aes(y = OS_MONTHS, x = AGE, color = SEX)) +
  geom_point()
```

## 1.6. Analysis of SARS-Cov2 vaccination data
```{r}
#### analyse the time series data on vaccination against SARS-Cov2
#### download at Impfdashboard
library(tidyr)
library(tidyverse)
library(dplyr)
library(ggplot2)

vac_deliveries  <- read_tsv("https://raw.githubusercontent.com/tlobnow/Resources/main/R_Introduction/CompBio_Data/germany_deliveries_timeseries_v2.tsv")
vac_by_state    <- read_tsv("https://raw.githubusercontent.com/tlobnow/Resources/main/R_Introduction/CompBio_Data/germany_vaccinations_by_state.tsv")
vac_timeseries  <- read_tsv("https://raw.githubusercontent.com/tlobnow/Resources/main/R_Introduction/CompBio_Data/germany_vaccinations_timeseries_v2.tsv")

glimpse(vac_deliveries)
glimpse(vac_by_state)
glimpse(vac_timeseries)

#### Deliveries over time
vac_deliveries %>%
  #filter(region == "DE-BB") %>%
  ggplot(aes(date, dosen, col = region)) + 
  geom_point()
```

```{r}
#### better to look at cumulative delivery over time:
#### define cum_dose

glimpse(vac_deliveries$region)
vac_deliveries %>%
  #filter(region == c("DE-BY", "DE-BB", "DE-BW")) %>%
  arrange(date) %>%
  group_by(region) %>%
  mutate(cum_doses = cumsum(dosen)) %>%
  ggplot(aes(date, cum_doses, col = region)) +
  geom_line()
```

```{r}
#### change code so that you plot types of vaccine delivered:
vac_deliveries %>%
  arrange(date) %>%
  group_by(impfstoff) %>%
  mutate(cum_doses = cumsum(dosen)) %>%
  ggplot(aes(date, cum_doses, col = impfstoff)) +
  geom_line()
```

```{r}
#### analyse delivered vaccine per state
#### first calculate delivery until today:

cumulative_deliveries_per_state <- vac_deliveries %>%
  group_by(region) %>%
  summarise(n = sum(dosen))

#### now merge with vac per state --> define vac_rate
vac_rate <- cumulative_deliveries_per_state %>%
  left_join(vac_by_state, by = c("region"="code")) %>%
  mutate(rate = vaccinationsTotal/n)
```

```{r}
#### plot as bar graph:
vac_rate %>%
  ggplot(aes(region, rate)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 90))
```

```{r}
#### plot number of vacs per time:

vac_timeseries %>%
  ggplot(aes(date, dosen_differenz_zum_vortag)) +
  geom_point() +
  geom_line()
```

```{r}
#### weekdays have differing rates
#### mark weekdays with color --> with wday() of lubridate

library(lubridate)
library(tidyverse)
library(ggplot2)

vac_timeseries %>%
  mutate(day_of_week = date %>% wday() %>% as_factor()) %>%
  ggplot(aes(date, dosen_differenz_zum_vortag, col = day_of_week)) +
  #geom_point() +
  geom_line() +
  facet_wrap(~day_of_week)
```