---
title:  "Correlation & Regression"
author: 'Finn Lobnow'
output: html_document
---

# Characterization of bivariate relationships
We look at four things:

1. Form       = linear, quadratic, non-linear
2. Direction  = positive, negative
3. Strength   = how much scatter/noise
4. Outliers   



# Correlation

- -1 < Correlation coefficient < +1
- Sign indicates direction
- Magnitude indicates strength

- beware of other than linear correlations! simple mathematical interpretation is not sufficient
- add visual analysis to ensure you don't miss non-linear relationships!

--> Pearson product-moment correlation 
(dep. on context there are other means of correlation measurements)

``` {r}
# Compute correlation
ncbirths %>%
  summarize(N = n(), r = cor(mage, weight))

# Compute correlation for all non-missing pairs
ncbirths %>%
  summarize(N = n(), r = cor(weeks, weight, use = "pairwise.complete.obs"))
```



# Anscombe Data Set
- great way to see the importance of visual interpretation!
- synthetic data sets with all same correlation coefficients, means, Standard Deviations BUT DIFFERENT REGRESSIONS


1. plot the data
2. calculate the correlation coefficient
3. (log transform to see potential incr. correlation)

``` {r}
# Run this and look at the plot
ggplot(data = mammals, aes(x = BodyWt, y = BrainWt)) +
  geom_point() + 
  scale_x_log10() + 
  scale_y_log10()

# Correlation among mammals, with and without log
mammals %>%
  summarize(N = n(), 
            r = cor(BodyWt, BrainWt), 
            r_log = cor(log(BodyWt), log(BrainWt)))
```


==> CORRELATION DOES NOT IMPLY CAUSATION!

``` {r}

# Create faceted scatterplot
# noise data set = random x,y in z sets
noise %>% ggplot(aes(
  x,y
)) +
geom_point() +
facet_wrap(~z)

# Compute correlations for each dataset
noise_summary <- noise %>%
  group_by(z) %>%
  summarize(N = n(), spurious_cor = cor(x, y))

# Isolate sets with correlations above 0.2 in absolute strength
noise_summary %>%
  filter(abs(spurious_cor) > 0.2)

```


# VISUALIZATION OF LINEAR MODELS
``` {r}
# Scatterplot with regression line
ggplot(data = bdims, aes(x = hgt, y = wgt)) + 
  geom_point() + 
  geom_smooth(method = 'lm', se = FALSE)
```

# UNDERSTANDING LINEAR MODELS
Generic statistical model: 

    Response = f(explanatory) + noise
    

  
``` {r}

```



